---
title: "Download and process COSMIC files"
author: "Jim Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    fig_caption: yes
---

```{r global_setup, include=FALSE}
knitr::opts_chunk$set(dpi=300, fig.pos="H", fig.width=8, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE);

rerun.all<-FALSE; 

source_url<-"sftp-cancer.sanger.ac.uk/cosmic/grch38/cosmic/v75";
file_home<-paste(Sys.getenv('RCHIVE_HOME'), 'data/variant/public/cosmic/src', sep='/');
file_out<-paste(Sys.getenv('RCHIVE_HOME'), 'data/variant/public/cosmic/r', sep='/');
genome<-'GRCh38'; 

library(VariantAnnotation); 
```

# File source and locations

  - All source files were downloaded from: **`r source_url`** through SFTP protocol
    - Be aware of the [terms and conditions](http://cancer.sanger.ac.uk/cancergenome/assets/COSMIC_academic_license_march2015.pdf) of using these files   
  - Downloaded files were unziped and saved to: **`r file_home`** for further processing
  - Processed files were archived to: **`r file_out`**

# Processing of individual files

_!Quoted description was copied from COSMIC_

## CosmicCodingMuts.vcf
  
```{r CosmicCodingMuts.vcf, include=TRUE}
fn<-paste(file_out, 'CosmicCodingMuts', sep='/'); 
if (rerun.all | !file.exists(paste(fn, '.rds', sep=''))) {
  GtUtility::loadVcf(paste(file_home, 'CosmicCodingMuts.vcf', sep='/'), fn, genome); 
  load(paste(fn, '.rdata', sep='')); 
  gr<-rowData(vcf); 
  elementMetadata(gr)<-cbind(elementMetadata(gr), info(vcf)); 
  saveRDS(gr, paste(fn, '.rds', sep='')); 
  file.remove(paste(fn, '.rdata', sep='')); 
} else vcf<-readRDS(paste(fn, '.rds', sep='')); 

cat("head(info(vcf))\n"); 
head(info(vcf)); 
```
  - "_VCF file of all coding mutations in cosmic_"
  - Make sure unzip the downloaded **.gz** file first

## cancer_gene_census.csv
```{r cancer_gene_census.csv, include=TRUE}
fn<-paste(file_out, 'cancer_gene_census.rds', sep='/'); 
tbl<-read.csv(paste(file_home, 'cancer_gene_census.csv', sep='/')); 
saveRDS(tbl, fn); 
```

```{r save_script, include=FALSE}
##############################################################################################################
tm<-strsplit(as.character(Sys.time()), ' ')[[1]][1];
fn0<-paste(Sys.getenv("RCHIVE_HOME"), '/source/script/update/variant/ProcessCosmic.Rmd', sep='');
fn1<-paste(Sys.getenv("RCHIVE_HOME"), '/source/script/update/variant/log/', tm, '_ProcessCosmic.Rmd' , sep='');
file.copy(fn0, fn1)

```
 _END OF DOCUMENT_